---
title: "Merging mixture components"
author: "Marc Comas-Cufí"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Merging mixture components}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

The most standard parametric approach in cluster analysis assumes data can be modelled by a finite mixture distribution. The approach has two steps; first, a finite mixture distribution with probability density function \[ f(\;\cdot\; ; \pi_1, \dots, \pi_k, \theta_1, \dots, \theta_k) = \pi_1   (\;\cdot\; ; \theta_1) + \dots + \pi_k f(\;\cdot\; ; \theta_k), \] with $\sum_{j=1}^k = \pi_j =1$ is fitted to a sample $ X$, obtaining estimates $\hat{\pi}_1, \dots, \hat{\pi}_k$ and $\hat{\theta}_1 \dots \hat{\theta}_k$. After the fitting process, each observation $ x$ is assigned to the finite mixture component $j$, $1\leq j \leq k$, with $\hat{\pi}_j f( x ; \hat{\theta}_j)$ maximum. 

We are going to work with the dataset ex4.1 used in Baudry et el. (2010) and available in package `mclust`. To fit a finite mixture of gaussian distributions we are going to use the same package. 

```{r}
library(ggplot2)
library(mclust)
data(Baudry_etal_2010_JCGS_examples)
qplot(data=ex4.1, X1, X2)
```

# Finite mixture fitting

Function `Mclust` allows us to fit a mixture function to a dataset.

```{r}
m <- Mclust(ex4.1)
summary(m)
```



```{r}
library(mixpack)
library(dplyr)

dens = dmixnorm_solution(ex4.1, solution = m)
(df <- lapply(1:6, function(i) 
  (m$parameters$pro[i] * dmixnorm_solution(ex4.1, solution = m, part=i)) %>%
    data.frame %>% {./dens} %>%
    setNames(sprintf('p%02d', i)) ) %>% bind_cols)
```

The posterior probabilities are also available in the object returned by function `Mclust`.

```{r}
m$z %>% head
```


```{r}
table(apply(df, 1, which.max), m$classification)
```

```{r}
xlimits = seq(-3, 11, 0.05)
ylimits = seq(-3, 8, 0.05)
cm0 = expand.grid(X1 = xlimits, X2 = ylimits) %>% tbl_df %>%
  mutate(z = dmixnorm_solution(., solution=m))
```

```{r}
ggplot() + 
  geom_point(data=ex4.1, aes(x=X1, y=X2),alpha=0.2) + 
  stat_contour(data = cm0, aes(x=X1, y=X2, z=z))
```


```{r}
partition = list(1,2,3,4,5,6)
CN6 = lapply(partition, function(part){
  expand.grid(X1 = xlimits, X2 = ylimits) %>%
    tbl_df %>%
    mutate(z = dmixnorm_solution(., m, part = part),
           id = sprintf('{%s}',paste(part, collapse=',')))
}) %>% bind_rows

ggplot() + 
  geom_point(data=ex4.1, aes(x=X1, y=X2),alpha=0.2) + 
  stat_contour(data = CN6, aes(x=X1, y=X2, z=z, col=id))
```

## Using a specific partition

```{r}
partition = list(c(1,6,2),c(3,4),5)
CN6 = lapply(partition, function(part){
  expand.grid(X1 = xlimits, X2 = ylimits) %>%
    mutate(z = dmixnorm_solution(., m, part = part),
           id = sprintf('{%s}',paste(part, collapse=',')))
}) %>% bind_rows

ggplot() + 
  geom_point(data=ex4.1, aes(x=X1, y=X2),alpha=0.2) + 
  stat_contour(data = CN6, aes(x=X1, y=X2, z=z, col=id))
```

